---
title: "Filtering Strategy for the Global Cancer Analysis Pipeline"
editor:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

probes_table_base <- file.path("..", "..", "resources", "tables", "probes")

carcinomas_probe_counts_path <- file.path(probes_table_base, "carcinomas_probe_counts.html")
blastomas_probe_counts_path  <- file.path(probes_table_base, "blastomas_probe_counts.html")
leukemias_probe_counts_path  <- file.path(probes_table_base, "leukemias_probe_counts.html")
lymphomas_probe_counts_path  <- file.path(probes_table_base, "lymphomas_probe_counts.html")

safe_include_html <- function(path) {
  if (file.exists(path)) {
    cat(readLines(path, warn = FALSE), sep = "\n")
  } else {
    cat("⚠️ Table not available.\n\n")
  }
}
```

# Filtering of Microarray Expression Data

This document describes the role of probe filtering in the global cancer analysis workflow.
It should be read together with the preprocessing and analysis pipeline overviews, because filtering in the analysis pipeline is the stage at which the high-dimensional expression matrix is converted into comparison-specific feature sets for downstream complexity, entropy, and biological interpretation tasks.

Latent-space modeling uses a related but distinct feature-selection strategy defined in the preprocessing pipeline, and therefore does not necessarily operate on the same comparison-specific feature sets described here.

------------------------------------------------------------------------

## Current Analysis Configuration

The current analysis configuration supports two comparison-level probe-selection regimes, with variance filtering available in two modes:

| Method / mode | Main parameters | Primary emphasis |
|---|---:|---|
| `limma` | `logfc_cutoff = 0.33`, `pval_cutoff = 0.05` | Differential-expression-oriented selection within each predefined normal-tumor comparison |
| `variance`, `top_n` mode | `analysis_top_n = 3000` | Fixed-count high-variance probe selection within each predefined normal-tumor comparison |
| `variance`, `threshold` mode | `var_threshold = 0.75` | Quantile-threshold high-variance probe selection within each predefined normal-tumor comparison |

In the current configuration, the analysis pipeline uses:

```r
filter_method <- "variance"
variance_selection_mode <- "top_n"
analysis_top_n <- 3000
```

This comparison-specific `analysis_top_n` setting is distinct from the global top-N feature-selection strategy used for latent-space export during preprocessing.
Both may use the value 3000, but they define different feature spaces: the analysis setting defines local, comparison-specific high-variance probe sets, whereas the latent-space workflow uses a global chip-level high-variance feature set.

The analysis pipeline applies probe selection in Stage 3, immediately after expression matrices and normal/tumor comparison maps have been built.
The resulting filtered-probe objects are saved as chip-specific RDS files:

| Output | Description |
|---|---|
| `output/global_cancer/RData/filtered_probes/filtered_probes_hu35ksuba.rds` | Filtered probes for `hu35ksuba` comparisons |
| `output/global_cancer/RData/filtered_probes/filtered_probes_hu6800.rds` | Filtered probes for `hu6800` comparisons |

These files then provide the feature subsets used by the pairwise complexity and entropy engines.

------------------------------------------------------------------------

## Why Filtering Is Required

Gene-expression microarray datasets are high-dimensional but sample-limited.
The Ramaswamy dataset contains many thousands of probes, while each tissue-specific normal/tumor comparison contains far fewer samples.
Without filtering, downstream calculations would be dominated by probes that contribute little stable information: low signal, low reproducibility, low contrast, inconsistent annotation, or noise-dominated variation.

Filtering is therefore not only a statistical preprocessing step.
In this project it has three distinct roles:

1. it defines the probe space used by traditional complexity and entropy measures;
2. it provides a comparison-specific feature space that can be contrasted with the global feature-selection strategy used for latent-space modeling;
3. it shapes the biological interpretation of complexity, entropy, and latent geometry by determining which genes and gene sets are eligible to contribute to the analysis.

These roles are related but not identical.
A filter that is appropriate for differential-expression interpretation is not necessarily identical to one that stabilizes geometric structure, and neither necessarily matches the feature-selection regime used for latent-space modeling.
For that reason, filtering should be documented as an analytic choice rather than treated as a purely mechanical cleanup step.

------------------------------------------------------------------------

## Filtering and Traditional Complexity/Entropy Measures

The legacy R-side analysis computes complexity and entropy measures on normal/tumor expression matrices after comparison-specific probe selection.
In this setting, filtering defines the empirical state space over which covariance structure, dimensionality, dispersion, and entropy-like quantities are estimated.

Measures such as effective rank, participation ratio, eigenvalue entropy, anisotropy, and related covariance-derived statistics are sensitive to both the number and identity of retained probes.
Including many uninformative or noise-dominated probes can inflate apparent dimensionality or obscure coherent structure.
Conversely, retaining only highly differential probes can make the analysis more focused but may bias the measured state space toward large normal/tumor contrasts.

For this reason, the pipeline supports two conceptually distinct probe-selection regimes:

| Filtering mode | Interpretation for complexity/entropy analysis |
|---|---|
| `limma` | Complexity and entropy are assessed within a contrast-enriched probe space defined by differential expression between normal and tumor samples. |
| `variance` | Complexity and entropy are assessed within a dispersion-enriched probe space defined by high-variance probes within each comparison. |

Results obtained under different filtering regimes should not be interpreted as interchangeable without explicitly considering the effect of the filtering choice.

------------------------------------------------------------------------

## Filtering and Latent Space

The latent-space workflow reduces high-dimensional gene-expression data to a lower-dimensional representation.
Filtering affects this workflow because the model can only learn from the feature set it receives.

In the current project, the latent-space workflow and the R-side complexity/entropy analysis generally operate on different feature universes.

The latent-space pipeline uses a fixed, global set of high-variance probes selected across all samples on a given chip.
In contrast, the R-side analysis uses comparison-specific probe selection based on predefined normal/tumor contrasts.

As a result, latent-space representations and complexity/entropy measures are computed in different feature spaces.
Direct numerical correspondence between these systems should therefore be interpreted cautiously.
Agreement may reflect shared biological structure, whereas disagreement may arise from biological, geometric, or filtering-related differences.

Latent-space representations should therefore be interpreted as embeddings of a filtered transcriptomic subspace rather than of the full gene-expression space.

------------------------------------------------------------------------

## Filtering and Biological Meaning

Filtering determines the biological vocabulary available for interpretation.
GO terms, KEGG pathways, and other gene-set summaries depend on which probes remain after filtering and on how those probes map to annotation resources.

A large retained-probe count in a normal/tumor comparison may indicate broad transcriptional reorganization, but it can also reflect sample size, platform coverage, variance structure, or threshold choice.
A small retained-probe count may indicate a more subtle tumor-normal contrast, but it can also reflect low power, heterogeneous samples, conservative thresholds, or platform limitations.

Biological interpretation should therefore be phrased cautiously.
Filtered probe sets support statements such as:

- the comparison shows a larger or smaller number of probes meeting the configured filter;
- the retained probes are enriched for, or annotated to, particular biological functions or pathways;
- complexity/entropy or latent-space patterns are being estimated within the retained feature space.

They do not by themselves prove that a tissue has globally gained or lost biological complexity.
Rather, they help define the measured transcriptomic subspace in which complexity, entropy, and latent geometry are evaluated.

------------------------------------------------------------------------

## Limma-Oriented Filtering

The `limma` option is differential-expression-oriented.
In the current configuration, it uses:

```r
filter_method <- "limma"
logfc_cutoff  <- 0.33
pval_cutoff   <- 0.05
```

In this mode, the retained probes are those that pass the configured normal/tumor contrast criteria.
This makes the downstream analysis strongly tied to normal/tumor transcriptional divergence.
It is a defensible choice when the primary question is how the structure of expression changes among probes that differ between normal and malignant tissue.

The corresponding interpretive caution is that the measured complexity and entropy are conditional on a differential-expression filter.
The results describe the structure of the selected contrast-associated probe set, not the full transcriptome.

------------------------------------------------------------------------

## Variance-Oriented Filtering

The `variance` option is structure-oriented.
In the current configuration, it is parameterized as:

```r
filter_method <- "variance"
variance_selection_mode <- "top_n"
analysis_top_n <- 3000
```

The variance filter supports two modes.
In `top_n` mode, a fixed number of high-variance probes is retained per comparison.
In `threshold` mode, probes are retained according to the configured variance threshold:

```r
var_threshold <- 0.75
```

Retained probes are selected based on variability within each comparison-specific matrix.
This can emphasize dominant axes of variation and support geometric interpretation.

The corresponding interpretive caution is that variance is not synonymous with biological relevance.
Highly variable probes may reflect tumor biology, tissue identity, cellular mixture, technical effects, or other sources of heterogeneity.
For that reason, variance-filtered results should be interpreted together with metadata, annotation, and comparison-specific biological context.

------------------------------------------------------------------------

## Dynamic Filtering of Gene Sets

In addition to probe-level filtering, the analysis workflow also uses gene-set-level filtering for GO, KEGG, MSigDB, or combined gene-set modes.
The current configuration includes:

```r
gene_set_mode   <- "ALL"
quantile_cutoff <- 0.75
min_probes      <- 5
```

Gene-set filtering is necessary because gene-set sizes are highly uneven.
Some pathways contain many probes, while many GO terms contain very few mapped probes.
Very small gene sets can produce unstable estimates of complexity, entropy, or enrichment-like summaries.
Very large gene sets can dominate interpretation because they represent broad biological categories rather than specific mechanisms.

The goal of dynamic filtering is therefore to retain gene sets that are large enough to support stable calculation while avoiding indiscriminate inclusion of sparse or poorly supported terms.

### KEGG pathways

| Statistic | hu35ksuba | hu6800 |
|---|---:|---:|
| Median | 18 | 41.5 |
| 3rd quartile | 36 | 77.5 |
| Mean | 28.4 | 55.9 |
| Max | 427 | 676 |

Most KEGG pathways contain enough probes to support pathway-level summaries, although very small or very large pathway mappings still require caution.

### GO Molecular Function terms

| Statistic | hu35ksuba | hu6800 |
|---|---:|---:|
| Median | 2 | 2 |
| 3rd quartile | 6 | 6 |
| Mean | ~12 | ~12 |
| Max | ~6400 | ~5300 |

Most GO Molecular Function terms are very small in probe-count terms.
This makes minimum-probe thresholds especially important for GO-level calculations.
Without such thresholds, many GO summaries would be based on too few probes to support stable interpretation.

------------------------------------------------------------------------

## Application in the Current Analysis Pipeline

Filtering is implemented as Stage 3 of the analysis workflow.
The driver script sources the grouped filtering wrapper and high-variance helper, then runs filtering separately for each supported chip.
During this stage, the pipeline validates `variance_selection_mode`, derives the `top_n` argument only when variance filtering is run in `top_n` mode, and passes the filtering method, variance-selection mode, and active parameters to the grouped filtering wrapper.
The resulting RDS objects include provenance metadata describing the filtering regime used to generate them.
The current chips are:

| Chip | GEO platform | Role |
|---|---|---|
| `hu35ksuba` | `GPL98` | One Affymetrix platform used in the Ramaswamy dataset |
| `hu6800` | `GPL80` | Second Affymetrix platform used in the Ramaswamy dataset |

The filtering stage is comparison-aware because it operates against predefined normal/tumor comparison maps rather than applying one universal probe list to all comparisons.
This is appropriate for the R-side pairwise framework, where each comparison represents a distinct empirical contrast.

The practical output of this stage is not a final biological claim.
It is a set of comparison-specific feature lists that define the input space for subsequent complexity, entropy, aggregation, summary, and reporting stages.

------------------------------------------------------------------------

## Filtered Probe Counts

The following tables summarize filtered probe counts by tumor category.
These tables are currently included from pre-rendered HTML resources.
That is acceptable for documenting the existing report structure, but their role in PDF rendering should be reviewed separately when the reporting interface is audited.

### Carcinomas

```{r carcinomas_table, results='asis'}
safe_include_html(here::here(carcinomas_probe_counts_path))
```

### Blastomas

```{r blastomas_table, results='asis'}
safe_include_html(here::here(blastomas_probe_counts_path))
```

### Leukemias

```{r leukemias_table, results='asis'}
safe_include_html(here::here(leukemias_probe_counts_path))
```

### Lymphomas

```{r lymphomas_table, results='asis'}
safe_include_html(here::here(lymphomas_probe_counts_path))
```

------------------------------------------------------------------------

## Interpreting Filtered Probe Counts

Filtered probe counts should be interpreted as properties of a configured analysis run.
They depend on the selected filtering mode, thresholds, platform, comparison map, and available annotation.

A comparison with many retained probes may reflect broad normal/tumor transcriptional divergence.
It may also reflect larger sample size, stronger statistical power, or platform-specific probe coverage.
A comparison with few retained probes may reflect a subtler tumor-normal contrast, heterogeneous samples, conservative thresholds, or technical constraints.

For hematologic malignancies, high retained-probe counts may be consistent with strong transcriptional programs related to lineage, proliferation, immune signaling, or differentiation state.
For epithelial tumors, lower overlap across carcinoma comparisons may reflect tissue-specific oncogenic programs and differences in cell of origin.
These interpretations are plausible but should remain conditional on the filtering method and platform.

------------------------------------------------------------------------

## Shared Probes Within Tumor Categories

Shared-probe counts summarize how many filtered probes recur across comparisons within a broad tumor category.
They can be useful for identifying repeated transcriptional signals within a category, but they should not be treated as direct measures of biological similarity without considering tissue composition, platform coverage, and comparison design.

| Tumor category | Number of shared probes (`hu35ksuba`) | Number of shared probes (`hu6800`) |
|---|---:|---:|
| Carcinomas | 0 | 0 |
| Blastomas | 51 | 47 |
| Leukemias | 354 | 328 |
| Lymphomas | 181 | NA |

: Number of filtered probes shared across a tumor category

High shared-probe counts suggest that related comparisons may share recurrent transcriptional features.
Low shared-probe counts suggest either greater biological divergence, stronger tissue specificity, reduced statistical overlap, or platform-dependent limitations.
Missing values should not be interpreted as absence of signal; they may reflect absent comparisons or incomplete platform coverage.

------------------------------------------------------------------------

## Relationship to the Analysis Pipeline Overview

This filtering document corresponds to the current analysis overview in the following way:

| Analysis overview section | Filtering document connection |
|---|---|
| Stage 2: Build expression matrix and comparison maps | Defines the normal/tumor comparison maps used by comparison-aware filtering |
| Stage 3: Run probe selection | Main implementation point for this document |
| Stage 4: Run pairwise complexity and entropy comparisons | Consumes the selected probe sets as feature spaces |
| Stage 5: Aggregate engine results | Aggregates outputs whose values depend on the selected probe and gene-set filters |
| Stage 6: Summarize comparison-level patterns | Summarizes complexity and entropy patterns conditional on filtering choices |
| Stage 7-8: Latent workflows | Use a separate global feature-selection regime defined in preprocessing; this may share the same top-N value but does not define the same feature space as `analysis_top_n` |

This alignment is important because the same word, "filtering," can refer to several related but distinct decisions: selecting probes for pairwise analysis, selecting highly variable features for latent modeling, and selecting gene sets with enough mapped probes for stable biological interpretation.

------------------------------------------------------------------------

The empirical structure of filtered probe sets across comparisons is examined in the Results section, where intersection patterns and shared-feature distributions are analyzed using UpSet visualizations.
