---
title: "Comparison Report"
author: "Ali Al-Timimi"
format: html
params:
  data_file: NULL
execute:
  eval: true
---

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(stringr)
library(glue)
library(purrr)
library(knitr)
library(kableExtra)

if (is.null(params$data_file) || !file.exists(params$data_file)) {
  stop("Please provide a valid 'data_file' parameter pointing to an existing RDS file")
}

comp_data <- readRDS(params$data_file)

comparison     <- comp_data$comparison
summaries_df   <- comp_data$summaries_df
complexity_df  <- comp_data$complexity_df
entropy_df     <- comp_data$entropy_df
res_35k        <- comp_data$res_35k
res_6800       <- comp_data$res_6800

count_35k <- comp_data$count_35k
count_6800 <- comp_data$count_6800

summary_all <- summaries_df %>% filter(mode == "ALL")
summary_go <- summaries_df %>% filter(mode == "GO")
summary_kegg <- summaries_df %>% filter(mode == "KEGG")
summary_msig <- summaries_df %>% filter(mode == "MSIG")

# Cleaned name for image file references
clean_name <- gsub("[^a-zA-Z0-9]", "_", tolower(comparison))

plot_base <- file.path("..", "..", "resources", "plots")
table_base <- file.path("..", "..", "resources", "tables")

probes_path <- file.path(plot_base, "probes", paste0(clean_name, "_probe_counts.png"))
go_complex_path <- file.path(plot_base, "go", paste0(clean_name, "_go_complexity.png"))
go_entropy_path <- file.path(plot_base, "go", paste0(clean_name, "_go_entropy.png"))
kegg_complex_path <- file.path(plot_base, "kegg", paste0(clean_name, "_kegg_complexity.png"))
kegg_entropy_path <- file.path(plot_base, "kegg", paste0(clean_name, "_kegg_entropy.png"))

go_complex_table_path <- file.path(table_base, paste0(clean_name, "_go_complexity.html"))
go_entropy_table_path <- file.path(table_base, paste0(clean_name, "_go_entropy.html"))
kegg_complex_table_path <- file.path(table_base, paste0(clean_name, "_kegg_complexity.html"))
kegg_entropy_table_path <- file.path(table_base, paste0(clean_name, "_kegg_entropy.html"))
msig_complex_table_path <- file.path(table_base, paste0(clean_name, "_msig_complexity.html"))
msig_entropy_table_path <- file.path(table_base, paste0(clean_name, "_msig_entropy.html"))

go_complex_clusters_table_path <- file.path(table_base, "clusters", "go", paste0(clean_name, "_go_clusters.html"))
go_entropy_clusters_table_path <- file.path(table_base, "clusters", "go", paste0(clean_name, "_go_entropy_clusters.html"))

# Are we producing these plots?
# msig_complex_path <- file.path(plot_base, "msig", paste0(clean_name, "_msig_complexity.png"))
# msig_entropy_path <- file.path(plot_base, "mish", paste0(clean_name, "_msig_entropy.png"))

safe_include_image <- function(path, width = "75%") {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(glue::glue("âš ď¸Ź Plot not available.\n\n"))
  }
}

safe_include_html <- function(path) {
  if (file.exists(path)) {
    cat(paste(readLines(path, warn = FALSE), collapse = "\n"))
  } else {
    cat(glue::glue("âš ď¸Ź Table not available.\n\n"))
  }
}

```

::: {.panel-tabset}

### `r comparison` Overview

::: {.panel-tabset}

#### Probes
> -   **Cancer Category:** `r summaries_df %>% filter(mode == "ALL") %>% pull(cancer_category) %>% unique()`
> -   **Chip hu35ksuba**: `r count_35k` filtered probes
> -   **Chip hu6800**: `r count_6800` filtered probes

```{r probe_image, echo=FALSE}
safe_include_image(probes_path)
```

#### Summary Assessment

**Overall Complexity & Entropy Assessment**

| Measure | All Filtered Probes | GO MF Terms | KEGG Pathways | HALLMARK Genes |
|---------|---------------------|-----------------------------|---------------|----------------|
| **Complexity** | `r summary_all$complexity_strength_label` (`r summary_all$complexity_confidence_label`) | `r summary_go$complexity_strength_label` (`r summary_go$complexity_confidence_label`) | `r summary_kegg$complexity_strength_label` (`r summary_kegg$complexity_confidence_label`) | `r summary_msig$complexity_strength_label` (`r summary_msig$complexity_confidence_label`) |
| **Shannon Entropy** | `r summary_all$strength_label_shannon` (`r summary_all$confidence_label_shannon`) | `r summary_go$strength_label_shannon` (`r summary_go$confidence_label_shannon`) | `r summary_kegg$strength_label_shannon` (`r summary_kegg$confidence_label_shannon`) | `r summary_msig$strength_label_shannon` (`r summary_msig$confidence_label_shannon`) |
| **Spectral Entropy** | `r summary_all$strength_label_spectral` (`r summary_all$confidence_label_spectral`) | `r summary_go$strength_label_spectral` (`r summary_go$confidence_label_spectral`) | `r summary_kegg$strength_label_spectral` (`r summary_kegg$confidence_label_spectral`) | `r summary_msig$strength_label_spectral` (`r summary_msig$confidence_label_spectral`) |

**Notes**
-   **Gain** = increased complexity from normal to tumor\
-   **Loss** = complexity reduction from normal to tumor\
-   **Chaotic** = increased entropy from normal to tumor\
-   **Anti-Chaotic** = entropy reduction from normal to tumor\
-   All p-values unadjusted unless noted

:::

### GO Biological Process Clusters

::: {.panel-tabset}

#### Complexity Change

```{r go_bp_complex_clusters, echo=FALSE, results='asis'}
safe_include_html(go_complex_clusters_table_path)
```

#### Spectral Entropy Change

```{r go_bp_entropy_clusters, echo=FALSE, results='asis'}
safe_include_html(go_entropy_clusters_table_path)
```

:::

### GO Molecular Function Terms

::: {.panel-tabset}

#### Complexity Change

**Significant GO Molecular Function Terms (Complexity)**

```{r go_complexity_plot, echo=FALSE}
safe_include_image(go_complex_path)
```

```{r go_mf_terms_complexity, echo=FALSE, results = 'asis'}
safe_include_html(go_complex_table_path)
```

#### Spectral Entropy Change

**Significant GO Molecular Function Terms (Spectral Entropy)**

```{r go_entropy_plot, echo=FALSE}
safe_include_image(go_entropy_path)
```

```{r go_mf_terms_entropy, echo=FALSE, results='asis'}
safe_include_html(go_entropy_table_path)
```

:::

### KEGG Pathway Analysis

::: {.panel-tabset}

#### Complexity Change

**Significant KEGG Pathways (Complexity)**

```{r kegg_complexity_plot, echo=FALSE}
safe_include_image(kegg_complex_path)
```

```{r kegg_pathways_complexity, echo=FALSE, results = 'asis'}
safe_include_html(kegg_complex_table_path)
```

#### Spectral Entropy Change

**Significant KEGG Pathways (Spectral Entropy)**

```{r kegg_entropy_plot, echo=FALSE}
safe_include_image(kegg_entropy_path)
```

```{r kegg_pathways_entropy, echo=FALSE, results='asis'}
safe_include_html(kegg_entropy_table_path)
```

:::

### HALLMARK Genes Analysis

::: {.panel-tabset}

#### Complexity Change

**Significant HALLMARK Genes (Complexity)**

```{r msig_pathways_complexity, echo=FALSE, results='asis'}
safe_include_html(msig_complex_table_path)
```

#### Spectral Entropy Change

**Significant HALLMARK Genes (Spectral Entropy)**

```{r msig_pathways_entropy, echo=FALSE, results='asis'}
safe_include_html(msig_entropy_table_path)
```

#### Latent Space Change


:::

:::