---
title: "Latent space geometry results"
format: html
editor: 
  markdown: 
    wrap: 72
---

## Latent Space Geometry of Cancer

### Setup

```{r}
library(tidyverse)
library(gt)

# csvs <- file.path("quarto", "resources", "tables", "probes")

# Paths (adjust if needed)
taxonomy_path <- "latent_taxonomy_table.csv"
summary_path  <- "latent_regime_summary_table.csv"
narrative_path <- "latent_regime_summary_narrative.csv"

taxonomy <- read_csv(taxonomy_path, show_col_types = FALSE)
regime_summary <- read_csv(summary_path, show_col_types = FALSE)
regime_narrative <- read_csv(narrative_path, show_col_types = FALSE)
```

------------------------------------------------------------------------

# Methods

## Latent Representation

Gene expression data were embedded into a low-dimensional latent space
using a variational autoencoder (VAE). Input features consisted of the
top 3000 probes selected by variance. The encoder maps each sample to a
latent vector, and all subsequent analyses are performed in this learned
space.

## Tissue-Specific Comparisons

Samples were grouped by tissue and disease state (normal vs tumor). For
each tissue, latent representations were partitioned into normal and
tumor groups, defining a tissue-specific transition.

## Geometric Measures

For each group, covariance matrices were computed and decomposed into
eigenvalues. From these, the following measures were derived:

-   **Participation Ratio (PR):** effective dimensionality

-   **Effective Rank (EffRank):** spectral uniformity

-   **Condition Number (κ):** axis dominance

-   **Anisotropy:** leading eigenvalue fraction

Spatial measures included centroid distance and group radius.

## Delta Metrics

For each metric MMM, changes were defined as:

ΔM=Mtumor−Mnormal\Delta M = M\_{\text{tumor}} -
M\_{\text{normal}}ΔM=Mtumor​−Mnormal​

## Taxonomy Construction

A geometric taxonomy was defined using the joint behavior of ΔEffRank
and ΔPR:

-   **Regime A:** anisotropic expansion (EffRank ↓, PR ↑)

-   **Regime B:** compression / canalization (EffRank ↑, PR ↓)

-   **Regime C:** uniform expansion (EffRank ↑, PR ↑)

-   **Regime D:** collapse / simplification (EffRank ↓, PR ↓)

Transformation types were defined using centroid distance and magnitude
of ΔEffRank.

------------------------------------------------------------------------

# Results

## Regime Summary

```{r}
regime_summary %>%
  gt() %>%
  tab_header(
    title = "Latent-Space Regime Summary",
    subtitle = "Counts and percentages of geometric transformation regimes"
  ) %>%
  fmt_number(columns = where(is.numeric), decimals = 2)
```

------------------------------------------------------------------------

## Narrative Summary

```{r}
regime_narrative %>%
  select(regime_code, regime_label, summary_phrase) %>%
  gt() %>%
  tab_header(
    title = "Narrative Summary of Regimes"
  )
```

------------------------------------------------------------------------

## Dominant Transformation Pattern

Across all comparisons, the dominant regime is **anisotropic
expansion**, characterized by increasing dimensional spread (PR) and
decreasing spectral uniformity (EffRank).

This indicates that tumor states expand into additional degrees of
freedom, but in a highly uneven manner.

------------------------------------------------------------------------

## Taxonomy Table (Detailed)

```{r}
taxonomy %>%
  select(
    comparison,
    regime_label,
    transformation_type,
    effrank_delta,
    pr_delta,
    centroid_distance,
    summary_phrase
  ) %>%
  arrange(regime_label, desc(centroid_distance)) %>%
  gt() %>%
  tab_header(
    title = "Latent Geometry Taxonomy Table",
    subtitle = "Per-comparison geometric classification and summary"
  ) %>%
  fmt_number(
    columns = c(effrank_delta, pr_delta, centroid_distance),
    decimals = 3
  )
```

------------------------------------------------------------------------

## Decoupling of Dimensionality and Spectral Structure

A moderate negative correlation (≈ −0.41) was observed between ΔEffRank
and ΔPR, indicating that increases in dimensionality are accompanied by
decreased spectral uniformity. This reflects anisotropic expansion
rather than uniform dispersion.

------------------------------------------------------------------------

## Spatial vs Structural Transformation

Centroid displacement and structural change were negatively correlated
(≈ −0.47), indicating that:

-    some cancers undergo **large spatial shifts with minimal
    restructuring**

-    others exhibit **substantial internal reorganization with limited
    movement**

These represent distinct transformation modes.

------------------------------------------------------------------------

## Secondary Geometric Effects

Changes in condition number (κ) and anisotropy were weakly correlated,
suggesting that axis dominance and global anisotropy represent partially
independent geometric features.

------------------------------------------------------------------------

## Summary

Tumor progression in latent space is characterized by:

-    dominant **anisotropic expansion**

-    **decoupling** of dimensionality and spectral structure

-    separation of **spatial movement** and **internal reorganization**

These findings support a geometric, multidimensional view of cancer
complexity rather than a scalar notion of disorder.
